---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---



## Loading and preprocessing the data
```{r echo = TRUE}
# The data is read from the zip file and saved into a variable after processing the CSV file
activitymonitoringdataconnection <- unz("activity.zip", "activity.csv")
activitymonitoringdata <- read.csv(activitymonitoringdataconnection, sep=",", head = TRUE)
```


## 1. What is mean total number of steps taken per day?
## 1.1 Total number of steps taken per day
```{r}
activitymonitoringdataflag <- cbind( activitymonitoringdata, !is.na(activitymonitoringdata$steps))
colnames(activitymonitoringdataflag)[4] <- "flag"
activitymonitoringdataexNA <- activitymonitoringdataflag[activitymonitoringdataflag$flag == TRUE,]
activitymonitoringdataexNAtoAgg <- cbind(data.frame(cbind(activitymonitoringdataexNA$steps)), 
                                         activitymonitoringdataexNA$date)
colnames(activitymonitoringdataexNAtoAgg) <- c("steps", "date")

activitymonitoringdataexNAsum <- aggregate(as.numeric(activitymonitoringdataexNAtoAgg$steps), 
                                           by = list(activitymonitoringdataexNAtoAgg[,"date"]), 
                                           FUN = sum, na.rm = TRUE)
colnames(activitymonitoringdataexNAsum) <- c("date", "Number of steps")
activitymonitoringdataexNAsum
```
## 1.2 Histogram of the total number of steps taken each day
```{r}
hist(activitymonitoringdataexNAsum$'Number of steps', 
      main = "Histogram of the total number of steps taken each Day",
      breaks = 20,
      xlab = "steps",
      xaxt = 'n')
axis(1, at = seq(0,22000, by = 1000), labels = seq(0,22000, by = 1000) )

```
## 1.3 Mean and Median of the total number of steps taken per day
```{r}
activitymonitoringdataexNAMean <- aggregate(activitymonitoringdataexNAtoAgg$steps, 
                                            by = list(activitymonitoringdataexNAtoAgg[,"date"]), 
                                            FUN =  mean, na.rm = TRUE)
colnames(activitymonitoringdataexNAMean) <- c("Date", "Mean of Steps")
activitymonitoringdataexNAMedian <- aggregate(activitymonitoringdataexNAtoAgg$steps, 
                                              by = list(activitymonitoringdataexNAtoAgg[,"date"]), 
                                              FUN = median, na.rm = TRUE)
colnames(activitymonitoringdataexNAMedian) <- c("Date", "Median of Steps")
activitymonitoringdataMeanMedian <- merge(activitymonitoringdataexNAMean,
                                          activitymonitoringdataexNAMedian, by = 'Date' )
colnames(activitymonitoringdataMeanMedian) <- c("Date", "Mean Steps", "Median Steps")
activitymonitoringdataMeanMedian
```

## 2. What is the average daily activity pattern?
## 2.1 Time series plot of the 5-minute interval (x-axis)
```{r}
activitymonitoringdataflag <- cbind( activitymonitoringdata, !is.na(activitymonitoringdata$steps))
colnames(activitymonitoringdataflag)[4] <- "flag"
activitymonitoringdataexNA <- activitymonitoringdataflag[activitymonitoringdataflag$flag == TRUE,]
activitymonitoringdataexNAtoAgg2 <- cbind(data.frame(cbind(activitymonitoringdataexNA$steps)), 
                                          activitymonitoringdataexNA$interval)
colnames(activitymonitoringdataexNAtoAgg2) <- c("steps", "interval")
activitymonitoringdataexNAavg2 <- aggregate(as.numeric(activitymonitoringdataexNAtoAgg2$steps), 
                                            by = list(activitymonitoringdataexNAtoAgg2[,"interval"]), 
                                            FUN = mean, na.rm = TRUE)
colnames(activitymonitoringdataexNAavg2) <- c("interval", "average_steps")

plot(activitymonitoringdataexNAavg2$interval,
     as.numeric(activitymonitoringdataexNAavg2$average_steps),
     type = 'l',
     xlab = "Minutes of the day",
     ylab = "Average steps",
     xaxt = 'n')
axis(1, at = seq(0,2400, by = 100), labels = seq(0,2400, by = 100))
```

## 2.2 Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
activitymonitoringdataexNAavg2order <- activitymonitoringdataexNAavg2[order(-activitymonitoringdataexNAavg2$average_steps),]
activitymonitoringdataexNAavg2order$interval[1]
```


## Imputing missing values  



## Are there differences in activity patterns between weekdays and weekends?
