---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---



## Loading and preprocessing the data
```{r echo = TRUE}
# The data is read from the zip file and saved into a variable after processing the CSV file
activitymonitoringdataconnection <- unz("activity.zip", "activity.csv")
activitymonitoringdata <- read.csv(activitymonitoringdataconnection, sep=",", head = TRUE)
```


## 1. What is mean total number of steps taken per day?
## 1.1 Total number of steps taken per day
```{r}
activitymonitoringdataflag <- cbind( activitymonitoringdata, !is.na(activitymonitoringdata$steps))
colnames(activitymonitoringdataflag)[4] <- "flag"
activitymonitoringdataexNA <- activitymonitoringdataflag[activitymonitoringdataflag$flag == TRUE,]
activitymonitoringdataexNAtoAgg <- cbind(data.frame(cbind(activitymonitoringdataexNA$steps)), 
                                         activitymonitoringdataexNA$date)
colnames(activitymonitoringdataexNAtoAgg) <- c("steps", "date")

activitymonitoringdataexNAsum <- aggregate(as.numeric(activitymonitoringdataexNAtoAgg$steps), 
                                           by = list(activitymonitoringdataexNAtoAgg[,"date"]), 
                                           FUN = sum, na.rm = TRUE)
colnames(activitymonitoringdataexNAsum) <- c("date", "Number of steps")
activitymonitoringdataexNAsum
```



## 1.2 Histogram of the total number of steps taken each day
```{r}
hist(activitymonitoringdataexNAsum$'Number of steps', 
      main = "Histogram of the total number of steps taken each Day",
      breaks = 20,
      xlab = "steps",
      xaxt = 'n')
axis(1, at = seq(0,22000, by = 1000), labels = seq(0,22000, by = 1000) )

```
## 1.3 Mean and Median of the total number of steps taken per day
```{r}
activitymonitoringdataexNAMean <- aggregate(activitymonitoringdataexNAtoAgg$steps, 
                                            by = list(activitymonitoringdataexNAtoAgg[,"date"]), 
                                            FUN =  mean, na.rm = TRUE)
colnames(activitymonitoringdataexNAMean) <- c("Date", "Mean of Steps")
activitymonitoringdataexNAMedian <- aggregate(activitymonitoringdataexNAtoAgg$steps, 
                                              by = list(activitymonitoringdataexNAtoAgg[,"date"]), 
                                              FUN = median, na.rm = TRUE)
colnames(activitymonitoringdataexNAMedian) <- c("Date", "Median of Steps")
activitymonitoringdataMeanMedian <- merge(activitymonitoringdataexNAMean,
                                          activitymonitoringdataexNAMedian, by = 'Date' )
colnames(activitymonitoringdataMeanMedian) <- c("Date", "Mean Steps", "Median Steps")
activitymonitoringdataMeanMedian
```

## 2. What is the average daily activity pattern?
## 2.1 Time series plot of the 5-minute interval (x-axis)
```{r}
activitymonitoringdataflag <- cbind( activitymonitoringdata, !is.na(activitymonitoringdata$steps))
colnames(activitymonitoringdataflag)[4] <- "flag"
activitymonitoringdataexNA <- activitymonitoringdataflag[activitymonitoringdataflag$flag == TRUE,]
activitymonitoringdataexNAtoAgg2 <- cbind(data.frame(cbind(activitymonitoringdataexNA$steps)), 
                                          activitymonitoringdataexNA$interval)
colnames(activitymonitoringdataexNAtoAgg2) <- c("steps", "interval")
activitymonitoringdataexNAavg2 <- aggregate(as.numeric(activitymonitoringdataexNAtoAgg2$steps), 
                                            by = list(activitymonitoringdataexNAtoAgg2[,"interval"]), 
                                            FUN = mean, na.rm = TRUE)
colnames(activitymonitoringdataexNAavg2) <- c("interval", "average_steps")

plot(activitymonitoringdataexNAavg2$interval,
     as.numeric(activitymonitoringdataexNAavg2$average_steps),
     type = 'l',
     xlab = "Minutes of the day",
     ylab = "Average steps",
     xaxt = 'n')
axis(1, at = seq(0,2400, by = 100), labels = seq(0,2400, by = 100))
```

## 2.2 Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
activitymonitoringdataexNAavg2order <- 
  activitymonitoringdataexNAavg2[order(-activitymonitoringdataexNAavg2$average_steps),]
activitymonitoringdataexNAavg2order$interval[1]
```


## 3. Imputing missing values  
## 3.1 Total number of missing values in the dataset 
```{r}
activitymonitoringdataflagNumNA <- activitymonitoringdataflag[activitymonitoringdataflag$flag == FALSE,]
nrow(activitymonitoringdataflagNumNA)
```

## 3.2 Strategy for filling in all of the missing values in the dataset
##  The Strategy explanation : From dataset without NA, the mean of the steps is calculated by Interval and Weekday. 
##  Then based on the Interval and Weekday the calculated mean is assigned to the NA values in the activity dataset.
```{r}
activitymonitoringdataWDaysNAflag <- cbind( activitymonitoringdata ,   
                                          weekdays(as.Date(activitymonitoringdata$date)), 
                                          is.na(activitymonitoringdata$steps))
colnames(activitymonitoringdataWDaysNAflag) <- c("steps", "date", "Interval", "Weekday", "NAflag")
activitymonitoringdataWDaysYesNA <- activitymonitoringdataWDaysNAflag[activitymonitoringdataWDaysNAflag$NAflag == TRUE,]
colnames(activitymonitoringdataWDaysYesNA)  <- c("steps", "date", "Interval", "Weekday", "NAflag")
activitymonitoringdataWDaysNoNA <- activitymonitoringdataWDaysNAflag[activitymonitoringdataWDaysNAflag$NAflag == FALSE,]
colnames(activitymonitoringdataWDaysNoNA)  <- c("steps", "date", "Interval", "Weekday", "NAflag")
strategytofillNA <- aggregate(as.numeric(activitymonitoringdataWDaysNoNA$steps), 
                                            by = list(activitymonitoringdataWDaysNoNA[,"interval"], activitymonitoringdataWDaysNoNA[,"Weekday"]), 
                                            FUN = mean, na.rm = TRUE)
colnames(strategytofillNA) <- c("Interval", "Weekday", "MeanStepstofillNA")

activitymonitoringdataWDaysYesNAlen <- nrow(activitymonitoringdataWDaysYesNA)
for(i in 1:activitymonitoringdataWDaysYesNAlen)
  {
  strategytofillNAtemp1 <- strategytofillNA[strategytofillNA$Interval == activitymonitoringdataWDaysYesNA[i,]$Interval, ] 
  strategytofillNAtemp2 <- strategytofillNAtemp1[strategytofillNAtemp1$Weekday == activitymonitoringdataWDaysYesNA[i,]$Weekday, ]
  activitymonitoringdataWDaysYesNA[i,]$steps <- strategytofillNAtemp2$MeanStepstofillNA
  }
head(activitymonitoringdataWDaysYesNA)
unique(is.na(activitymonitoringdataWDaysYesNA$steps))
```

## 3.3 New dataset that is equal to the original dataset but with the missing data filled in.

```{r}
activitymonitorNewDataset <- rbind(activitymonitoringdataWDaysYesNA, activitymonitoringdataWDaysNoNA )
head(activitymonitorNewDataset)
```

## 3.4 Make a histogram of the total number of steps taken each day.
##     Calculate and report the mean and median total number of steps taken per day. 
##     Do these values differ from the estimates from the first part of the assignment? 
##     What is the impact of imputing missing data on the estimates of the total daily number of steps?

## Total number for steps taken each day.
```{r}
activitymonitorNewDatasetToAgg <- cbind(data.frame(cbind(activitymonitorNewDataset$steps)), 
                                         activitymonitorNewDataset$date)
colnames(activitymonitorNewDatasetToAgg) <- c("steps", "date")

activitymonitorNewDatasetSum <- aggregate(as.numeric(activitymonitorNewDatasetToAgg$steps), 
                                           by = list(activitymonitorNewDatasetToAgg[,"date"]), 
                                           FUN = sum, na.rm = TRUE)
colnames(activitymonitorNewDatasetSum) <- c("date", "Number of steps")


activitymonitorNewDatasetSum
```
## Histogram of the total number of steps taken each day
```{r}
hist(activitymonitorNewDatasetSum$'Number of steps', 
      main = "Histogram of the total number of steps taken each Day",
      breaks = 20,
      xlab = "steps",
      xaxt = 'n')
axis(1, at = seq(0,22000, by = 1000), labels = seq(0,22000, by = 1000) )
```


## New Dataset Mean and Median for the step taken per day
```{r}
## Mean and Median for the step taken per day.
activitymonitorNewDatasetMean <- aggregate(as.numeric(activitymonitorNewDatasetToAgg$steps), 
                                           by = list(activitymonitorNewDatasetToAgg[,"date"]), 
                                           FUN = mean, na.rm = TRUE)
colnames(activitymonitorNewDatasetMean) <- c("date", "Mean steps")
activitymonitorNewDatasetMedian <- aggregate(as.numeric(activitymonitorNewDatasetToAgg$steps), 
                                           by = list(activitymonitorNewDatasetToAgg[,"date"]), 
                                           FUN = median, na.rm = TRUE)
colnames(activitymonitorNewDatasetMedian) <- c("date", "Median steps")
activitymonitorNewDatasetMeanMedian <- merge(activitymonitorNewDatasetMean,
                                          activitymonitorNewDatasetMedian, by = 'date' )
colnames(activitymonitorNewDatasetMeanMedian) <- c("Date", "Mean Steps", "Median Steps")
activitymonitorNewDatasetMeanMedian
```

## Variance between the first part and the second part
## Impact Total Number of steps
```{r}
totalnumberofStepsImpacttemp <- merge (activitymonitoringdataexNAsum, activitymonitorNewDatasetSum, by = "date" , all = TRUE )
colnames(totalnumberofStepsImpacttemp) <- c("date", "OldTotalSteps" , "NewTotalSteps")
totalnumberofStepsImpact <- data.frame(totalnumberofStepsImpacttemp$date, 
                                       totalnumberofStepsImpacttemp$OldTotalSteps, 
                                       totalnumberofStepsImpacttemp$NewTotalSteps,
                                       totalnumberofStepsImpacttemp$OldTotalSteps - totalnumberofStepsImpacttemp$NewTotalSteps )
colnames(totalnumberofStepsImpact) <- c("date", "OldTotalSteps" , "NewTotalSteps","TotalstepsImpact")

totalnumberofStepsImpact
```


## Impact on Mean and Median steps
```{r}
MeanMedianStepsImpacttemp <- merge (activitymonitoringdataMeanMedian , activitymonitorNewDatasetMeanMedian,   by = "Date" , all = TRUE )
colnames(MeanMedianStepsImpacttemp) <- c("date", "OldMeanSteps", "OldMedianSteps" , "NewMeanSteps", "NewMedianSteps")
MeanMedianStepsImpact <- data.frame(MeanMedianStepsImpacttemp$date, 
                               MeanMedianStepsImpacttemp$OldMeanSteps,
                               MeanMedianStepsImpacttemp$NewMeanSteps,
                               MeanMedianStepsImpacttemp$OldMeanSteps - MeanMedianStepsImpacttemp$NewMeanSteps,
                               MeanMedianStepsImpacttemp$OldMedianSteps,
                               MeanMedianStepsImpacttemp$NewMedianSteps,
                               MeanMedianStepsImpacttemp$OldMedianSteps - MeanMedianStepsImpacttemp$NewMedianSteps )
colnames(MeanMedianStepsImpact) <- c("date", "OldMeanSteps" , "NewMeanSteps","MeanstepsImpact", "OldMedianSteps", "NewMedianSteps", "MedianstepsImpact")

MeanMedianStepsImpact

```



## Are there differences in activity patterns between weekdays and weekends?

```{r}
plot(activitymonitoringdataexNAavg2$interval,
     as.numeric(activitymonitoringdataexNAavg2$average_steps),
     type = 'l',
     xlab = "Minutes of the day",
     ylab = "Average steps",
     xaxt = 'n')
axis(1, at = seq(0,2400, by = 100), labels = seq(0,2400, by = 100))
```
